---
title: " PPOL 646 Final Project: SNAP / Food Stamp Recipients"
author: "Bill Radovich"
date: "5/17/2021"
output: html_document
---

<br><br><br>

<b>1.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(usmap)
library(rvest)
require(janitor)



theme_bill <- function() {
  theme_bw(base_size=12, base_family="Avenir") %+replace%
    theme(
      panel.background  = element_blank(),
      plot.background = element_rect(fill="grey98", colour=NA),
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      legend.position = "bottom"
    )
}
# ggsave("themes_bill.pdf", device=cairo_pdf, width=6, height=6)


# main data source: https://www.kaggle.com/jpmiller/publicassistance
SNAP_hist <- read.csv("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/SNAP_history_1969_2019.csv", stringsAsFactors=FALSE)
SNAP_hist[14,1] <- "1982"
SNAP_hist <- SNAP_hist %>%
  dplyr::rename(Year=Fiscal.Year)
SNAP_hist$Year <- as.Date(SNAP_hist$Year, format = "%Y")


inflat <- read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23e1e9f0&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=748&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=FPCPITOTLZGUSA&scale=left&cosd=1960-01-01&coed=2019-01-01&line_color=%234572a7&link_values=false&line_style=solid&mark_type=none&mw=3&lw=2&ost=-99999&oet=99999&mma=0&fml=a&fq=Annual&fam=avg&fgst=lin&fgsnd=2009-06-01&line_index=1&transformation=lin&vintage_date=2021-04-28&revision_date=2021-04-28&nd=1960-01-01")
inflat$DATE <- substring(inflat$DATE, 1, 4)
inflat <- dplyr::filter(inflat, DATE > "1968")
inflat <- inflat %>%
  dplyr::rename(Year=DATE, Annual_Rate=FPCPITOTLZGUSA)
inflat$Year <- as.Date(inflat$Year, format = "%Y")

inflat$BackCalc <- NA
nYears <- length(inflat$Year)
d <- 1 # 2020 dollar index value

for (i in 1:nYears){
  d <- d * 100
  d <- d / (100 + inflat$Annual_Rate[nYears-i+1])
  inflat$BackCalc[nYears-i+1] <- d
}

df1 <- full_join(SNAP_hist,inflat, by="Year")
df1$Average.Benefit.Per.Person_adj <- df1$Average.Benefit.Per.Person / df1$BackCalc



# 1 - SNAP benefit over time
ggplot() +
  geom_point(data=df1, aes(x=Year, y=Average.Benefit.Per.Person_adj), size=2, alpha=0.7, color="darkblue") +
  geom_line(data=df1, aes(x=Year, y=Average.Benefit.Per.Person_adj), size=1, color="darkblue") +
  ggtitle("Average SNAP Benefit per Person, 1969-2019", subtitle = "Sources: US Census Bureau, USDA, World Bank") +
  xlab("Fiscal Year") +
  ylab("Avg. Monthly Benefit / Person (2020 USD)") +
  scale_y_continuous(labels=scales::dollar_format(), limits=c(0, 160)) +
  annotate("segment", x = as.Date('1974-7-1'), xend = as.Date('1974-7-1'), y = 0, yend = 160, colour = "grey30", alpha = 0.8) +
  annotate("label", x = as.Date('1974-7-1'), y = 30, label = "1974: \nLinda Taylor trial") +
  annotate("rect", xmin = as.Date('1981-7-1'), xmax = as.Date('1982-7-1'), ymin = 0, ymax = 160, colour = "grey30", alpha = 0.2) +
  annotate("label", x = as.Date('1982-1-1'), y = 70, label = "1981-82: \nReagan-era \nwelfare reform") +
  annotate("segment", x = as.Date('1996-7-1'), xend = as.Date('1996-7-1'), y = 0, yend = 160, colour = "grey30", alpha = 0.8) +
  annotate("label", x = as.Date('1996-7-1'), y = 70, label = "1996: \nClinton-era \nwelfare reform") +
  annotate("segment", x = as.Date('2008-7-1'), xend = as.Date('2008-7-1'), y = 0, yend = 160, colour = "grey30", alpha = 0.8) +
  annotate("label", x = as.Date('2008-7-1'), y = 70, label = "2008: \nFood Stamp \nProgram changed \nto 'SNAP'") +
  theme_bill()

```

<br>
This graphic examines the lifetime of the Food Stamp Program / SNAP benefits granted to participants over the last 50 years. Significant events in the battle over welfare during this time period are highlighted. The oft-hurled stereotype of the "Welfare Queen" was born out of the publicised and nationally televised trial of Linda Taylor in 1974 - a political straw man that a common welfare recipient was a lazy black woman "gaming the system" for money - used by Ronald Reagan and anti-welfare advocates alike to pass new restrictions on welfare during the first years of his presidency - including Food Stamps. Following a hunger crisis in the mid-to-late-80s and a brief increase in benefits to recipients, welfare was reformed once again during the Clinton administration in 1996 - again heavily fueled by the "welfare queen" stereotype held by lawmakers. During the 2008-09 Global Recession, the Food Stamp program was heavily modified into the 'SNAP' program and replaced references to "stamps" or "coupon" with "card" or "EBT" to eradicate semantics that were previously used in racist attacks on the program and its recipients. SNAP funding has decreased in the past decade with a Republican-controlled congress influenced by the anti-welfare Tea Party movement and deficit hawks.
<br><br><br>
<b>2.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Children <- read.csv("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/WICAgencies2016ytd/Children_Participating.csv")
keeps1 <- c("State.Agency.or.Indian.Tribal.Organization","Average.Participation")
Children <- Children[keeps1]
Children <- Children %>%
  dplyr::rename(Name=State.Agency.or.Indian.Tribal.Organization)


Food_Costs <- read.csv("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/WICAgencies2016ytd/Food_Costs.csv")
keeps2 <- c("State.Agency.or.Indian.Tribal.Organization","Cumulative.Cost")
Food_Costs <- Food_Costs[keeps2]
Food_Costs <- Food_Costs %>%
  dplyr::rename(Name=State.Agency.or.Indian.Tribal.Organization)


est16us <- read.csv("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/est13_16us/est16us.csv", stringsAsFactors=FALSE)
labs <- est16us[1,]
est16us <- est16us %>% janitor::row_to_names(row_number = 1)
keeps3 <- c("State FIPS Code", "Postal Code", "Name", "Poverty Estimate, All Ages", "Poverty Percent, All Ages", "Poverty Estimate, Age 0-17", "Poverty Percent, Age 0-17", "Median Household Income", "Poverty Estimate, Age 0-4", "Poverty Percent, Age 0-4")
est16us_1 <- est16us[keeps3]
est16us_1 <- est16us_1 %>%
  dplyr::rename(PovPctAllAges="Poverty Percent, All Ages",
                PovPct0_17="Poverty Percent, Age 0-17",
                PovPct0_4="Poverty Percent, Age 0-4",
                PC="Postal Code",
                MedInc="Median Household Income")


df2 <- inner_join(Children, Food_Costs, by="Name")
df2 <- na.omit(df2)
df2 <- inner_join(est16us_1, df2, by="Name")
df2$PovPctAllAges <- as.numeric(as.character(df2$PovPctAllAges))
df2$PovPct0_17 <- as.numeric(as.character(df2$PovPct0_17))
df2$PovPct0_4 <- as.numeric(as.character(df2$PovPct0_4))
df2$MedInc <- as.numeric(as.character(df2$MedInc))
df2 <- df2 %>%
  dplyr::rename(fips="State FIPS Code")

ll <- min(df2$PovPctAllAges)
ul <- max(df2$PovPct0_4)



# 2 - Poverty by State, all ages
plot_usmap(regions = "states", data = df2, values = "PovPctAllAges") +
  scale_fill_gradient(name="Poverty %, All Ages (2016)", low="#99FF99", high="#990000", limits=c(ll, ul)) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(caption = "- US Definition of Poverty: an individual with income less than $36 per day or a family of four with income less than $72 per day", size = 2.5, colour = "black")

```
<br><br>
The distribution of poverty across the United States is shown in the map above. The light green highlighting states in the Great Plains, New England, the Mid-Atlantic, Alaska, and Hawaii indicate these regions generally have a lower poverty rate hovering around 8-12%. Darker orange-highlighted states in Appalachia and the Sun Belt have poverty rates of up to 20% - including states like Mississippi, New Mexico, West Virginia, and Kentucky. The West Coast and Midwest are somewhere in-between. The goal of this map was to gain insight into the actual poverty rates geographically and how these relate to the histories of each state and/or region of the country.
<br><br><br>
<b>3.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# 3 - Poverty by state, ages 0-4
plot_usmap(regions = "states", data = df2, values = "PovPct0_4") +
  scale_fill_gradient(name="Poverty %, Ages 0-4 (2016)", low="#99FF99", high="#990000", limits=c(ll, ul)) +
  theme_void() +
  theme(legend.position = "bottom") +
  labs(caption = "- US Definition of Poverty: an individual with income less than $36 per day or a family of four with income less than $72 per day", size = 2.5, colour = "black")

```
<br><br>
This map displays the poverty rate for children ages 0 to 4 on the same scale as the one above. It is clear that the distribution of poverty generally remains the same across the aforementioned regions and states. In states like New Hampshire, Utah, Virginia, Wyoming, and Maryland, the poverty rate for children 0-4 still remains just above 10%. However, the same Sun Belt and Appalachian states have significantly higher rates of poverty among young children - above 30% in Mississippi, New Mexico, and Louisiana. Despite the attacks on mothers on welfare for decades, babies are more likely than adults to be in poverty and reliant on public assistance.
<br><br><br>
<b>4.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
Avg_Cost <- read.csv("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/WICAgencies2016ytd/Average_Food_Cost_Per_Person.csv", stringsAsFactors=FALSE)
keeps4 <- c("State.Agency.or.Indian.Tribal.Organization", "Cumulative.Average")
Avg_Cost <- Avg_Cost[keeps4]
Avg_Cost <- Avg_Cost %>%
  dplyr::rename(Name=State.Agency.or.Indian.Tribal.Organization)

df2 <- inner_join(df2, Avg_Cost, by="Name")
df2 <- na.omit(df2)

gini <- read.csv("https://datacenter.prb.org/download/usdata/indicator/gini/csv", skip=3)
gini <- gini %>%
  dplyr::filter(TimeFrame == "2014-2018")

df3 <- inner_join(df2, gini, by="Name")
df3 <- df3 %>%
  dplyr::rename(Gini="Data")



# 4 - Median Income vs. Average SNAP Benefit per Child by State
ggplot() +
  geom_point(data=df3, aes(x=MedInc, y=Cumulative.Average, color=Gini), size=3, alpha=0.8) +
  geom_text(data=df3, aes(x=MedInc, y=Cumulative.Average, label=ifelse(Gini == max(Gini) | Gini == min(Gini), as.character(Name),'')), vjust=1.5) +
  scale_colour_gradient(low = "lightblue", high = "#443010") +
  xlab("Median Income (USD)") +
  ylab("Average SNAP Cost per Child (USD)") +
  scale_x_continuous(labels=scales::dollar_format()) +
  scale_y_continuous(labels=scales::dollar_format()) +
  ggtitle("SNAP Cost per Child vs. Median Income by State (2016)", subtitle = "Sources: US Census Bureau, USDA, datacenter.prb.org") +
  theme_bill()

```
<br><br>
The average monthly SNAP cost per child vs. medium income by state shows that there is no real correlation between the two measures. Low-income states could reasonably have high SNAP costs for children because of the massive poverty rates (Mississippi, top-right corner), or low SNAP costs due to lower cost of living. The same is true for high-income states - New Hampshire ($70,986/year median income) has low SNAP costs - D.C. has moderate per child SNAP costs - and Alaska's are among the highest in the country. This can be explained by New Hampshire's low wealth inequality and D.C.'s high wealth inequality (measured by Gini index). Alaska, despite having the lowest Gini in the U.S., can have its high SNAP costs attributed to the cost of importing basic foods and goods into the state. The costliness to feed children on SNAP is likely influenced more by historical and/or geographic factors of the region, wealth inequality, and cost of living rather than a state's income.
<br><br><br>
<b>5.)</b>

<br><br>
This interactive map aims to show the proportion of SNAP recipients whom are children, or women who are pregnant or postpartum. In most states, children make up more than half of recipients. Adult women make up around 20-25% of total SNAP recipients - including adult women who are pregnant or postpartum, who alone make up around 14%-20% depending on state. In Puerto Rico, 63.3% of recipients are children.
<br><br><br>
<b>6.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ggrepel)


df6 <- read.csv("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/df6.csv")
pop16_50 <- sum(df6$POPESTIMATE2016)


TotParticip <- read.csv("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/WICAgencies2016ytd/Total_Number_of_Participants.csv")
TotParticip <- TotParticip[complete.cases(TotParticip), ]
keeps5 <- c("State.Agency.or.Indian.Tribal.Organization","Average.Participation")
TotParticip <- TotParticip[keeps5]
TotParticip <- TotParticip %>%
  dplyr::rename(Name=State.Agency.or.Indian.Tribal.Organization)


# source: http://navajotimes.com/news/2012/0112/012612census.php
reservation_percentage <- 0.22
# source: https://www.census.gov/newsroom/facts-for-features/2017/aian-month.html#:~:text=6.7%20million,the%20total%20population%20in%202016.
# native population living on reservations (approx.)
nativepop16 <- 6700000 * reservation_percentage


# source: https://www.ncsl.org/Portals/1/Documents/magazine/articles/2018/SL_0118-Stats.pdf
territoriespop16 <- 3789851
territories <- c("Guam", "American Samoa", "Northern Marianas", "Puerto Rico", "Virgin Islands")
territoriesParticip <- TotParticip[TotParticip$Name %in% territories,]
territoriesParticip <- sum(territoriesParticip$Average.Participation)


states <- df6$STNAME
# Mountain Plains is part of a SNAP-Ed program to get children and adults in mountain region states to eat healthier
# includes some American Indians living on reservations
# I estimated this using the 3% American Indian participation times 22% of overall American Indians that live on reservations
# I account for this after transforming the data
# source: https://community-nutrition-education.extension.org/wp-content/uploads/2020/09/MPR-Impact-Report_Print-Ready.pdf
levels(states) <- c(levels(states),as.factor("Mountain Plains"))
nativeParticip <- TotParticip[TotParticip$Name %in% territories == FALSE & TotParticip$Name %in% states == FALSE,]
nativeParticip <- nativeParticip %>%
  dplyr::filter(Name != "Mountain Plains")
nativeParticip <- sum(nativeParticip$Average.Participation)


TotParticip <- TotParticip[complete.cases(TotParticip), ]
statesParticip <- TotParticip %>%
  dplyr::filter(Name %in% states)
statesParticip <- sum(statesParticip$Average.Participation)


MtnPlns <- TotParticip %>%
  dplyr::filter(Name == "Mountain Plains")
MtnPlnsParticip <- as.numeric(MtnPlns$Average.Participation)


totPop16 <- pop16_50 + territoriespop16 + nativepop16
totParticip16 <- statesParticip + territoriesParticip + nativeParticip + MtnPlnsParticip


MtnPlnsUS <- MtnPlnsParticip * (1 - reservation_percentage) * (1 - 0.03)
MtnPlnsRes <- MtnPlnsParticip - MtnPlnsUS


df7 <- data.frame(
  category=c("50 States + DC", "US Territories", "American Indian Reservations"),
  Population=c(pop16_50/totPop16, territoriespop16/totPop16, nativepop16/totPop16),
  Population_label=c(scales::label_percent(accuracy=0.001)(pop16_50/totPop16), scales::label_percent(accuracy=0.001)(territoriespop16/totPop16), scales::label_percent(accuracy=0.001)(nativepop16/totPop16)),
  Participation=c((statesParticip+MtnPlnsUS)/totParticip16, territoriesParticip/totParticip16, (nativeParticip+MtnPlnsRes)/totParticip16),
  Participation_label=c(scales::label_percent(accuracy=0.001)((statesParticip+MtnPlnsUS)/totParticip16), scales::label_percent(accuracy=0.001)(territoriesParticip/totParticip16), scales::label_percent(accuracy=0.001)((nativeParticip+MtnPlnsRes)/totParticip16))
  #Rate=c(scales::label_percent(accuracy=0.001)(statesParticip/pop16_50), scales::label_percent(accuracy=0.001)(territoriesParticip/territoriespop16), scales::label_percent(accuracy=0.001)(nativeParticip/nativepop16))
)


require(reshape)
df7 <- reshape::melt(df7, id=c("category", "Population_label", "Participation_label"))
for (i in 1:length(df7$Population_label)){
  if (df7$variable[i] != "Population"){
    df7$Population_label[i] <- " "
  }
}
for (i in 1:length(df7$Participation_label)){
  if (df7$variable[i] != "Participation"){
    df7$Participation_label[i] <- " "
  }
}
df7$label <- rep(0, length(df7$variable))
for (i in 1:length(df7$label)){
  if (is.na(df7$Population_label[i])){
    df7$label[i] <- as.character(df7$Participation_label[i])
  }
  if (is.na(df7$Participation_label[i])){
    df7$label[i] <- as.character(df7$Population_label[i])
  }
}
df7 <- df7[c("category", "variable", "value", "label")]



# 6 - SNAP Participation in 50 States + DC vs. US Territories vs. Indian Reservations
ggplot(df7, aes(
  x=variable,
  y=value,
  group=category,
  label=label
)) +
  geom_col(aes(fill=category)) +
  coord_flip() +
  coord_polar("y") +
  scale_y_continuous(labels=scales::percent_format()) +
  ggtitle("SNAP Participation Makeup (Outer) vs. \nUS Population Makeup (Inner) - 2016", subtitle = "Sources: US Census Bureau, USDA, navajotimes.com, \nwww.ncsl.org, community-nutrition-education.extension.org") +
  theme_bill() +
  ggrepel::geom_label_repel(box.padding = 1, label.padding = 0.3, label.size = 0.3, point.padding = .1, color="#111111", alpha=0.6, position = position_stack()) +
  theme(axis.title.x=element_blank(),
        axis.title.y=element_blank()) +
  labs(caption = "- US Territories: Puerto Rico, Guam, Virgin Islands, \nAmerican Samoa, and Northern Marianas.", size = 2.5, colour = "black")

```
<br><br>
This donut chart highlights the massive discrepancy between the proportion of Americans outside the 50 states + DC in US Territories and American Indian Reservations and the much more significant proportion those Americans are on SNAP. American Indians living on Reservations, despite making up less than (an estimated) 0.449% of the US population in 2016, they made up an estimated 2.138% of all SNAP participants/recipients. Those living in the 5 US Territories - Guam, the Virgin Islands, American Samoa, Puerto Rico, and Northern Marianas - made up an estimated 2.154% of welfare recipients despite having 1.155% of the US population.
<br><br><br>
<b>7.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(dplyr)
library(stringr)
# source: https://www.ers.usda.gov/media/10674/trends.xlsx
df8 <- read_excel("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/trends.xlsx", sheet=1, col_names=TRUE, skip=1)
df8 <- as.data.frame(na.omit(df8))
names(df8) <- str_replace_all(names(df8), c(" " = ""))
df8 <- df8 %>%
  dplyr::rename(Food_Insecure='Foodinsecurity(includeslowandverylowfoodsecurity)',
         Very_Insecure=Verylowfoodsecurity)
df8$Year <- as.Date(df8$Year, format = "%Y")


df1$Year <- as.Date(df1$Year, format = "%Y")
df8.1 <- inner_join(df1, df8, by="Year")
df8.1$Year <- substr(df8.1$Year, 1, 4)
df8.1$Food_Insecure <- as.numeric(df8.1$Food_Insecure)/100
df8.1$Very_Insecure <- as.numeric(df8.1$Very_Insecure)/100



# 7 - Hunger vs. SNAP Funding, 1995-2019
ggplot(data=df8.1, aes(x=Average.Benefit.Per.Person_adj, y=Food_Insecure, label=ifelse(as.numeric(Year) %% 5 == 0 | Year == max(Year), as.character(Year), ""))) +
  geom_point(size=4, color="red", alpha=0.4) +
  geom_segment(data=df8.1, color="red",
                aes(
                    xend=c(tail(Average.Benefit.Per.Person_adj, n=-1), NA),
                    yend=c(tail(Food_Insecure, n=-1), NA)
                  ),
                  arrow=arrow(length=unit(0.4, "cm"))
      ) +
  annotate("rect", xmin = 122, xmax = 150, ymin = .1445, ymax = .148, colour = "grey30", alpha = 0.2) +
  annotate("label", x = 118, y = .145, label = "2008-09: \nGreat Recession") +
  geom_text_repel(box.padding=2, alpha=0.5) +
  theme_bill() +
  scale_x_continuous(labels=scales::dollar_format()) +
  scale_y_continuous(labels=scales::percent_format()) +
  xlab("Average Monthly Benefit per Person (2020 USD)") +
  ylab("% of Americans Who are Food Insecure") +
  ggtitle("Food Insecurity vs. Average SNAP Benefit per Person, 1995-2019", subtitle="Sources: US Census Bureau, USDA") +
  labs(caption = "- The USDA defines food insecurity as a lack of consistent access to enough food for an active, healthy life.", size = 2.5, colour = "black")

```
<br><br>
This graph follows the relationship between the % of Americans suffering from hunger, or food insecurity, vs. the average monthly Food Stamp / SNAP benefit per participant. Following over a decade of the Clinton and Bush Jr. administrations maintaining a relatively modest average Food Stamp benefit per participant, the Great Recession caused the worst economic crisis since 1929, and a resulting spike in hunger. Highlighted on the graph are the two data points during the Recession, where hunger remained relatively high until several years of increased SNAP benefits to Americans. This quickly fell back down after 2010, closer to where it was before the Recession.
<br><br><br>
<b>8.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
df9 <- df8.1 %>%
  dplyr::filter(Year == 2016)
df9 <- df9[c("Total.Benefits.M.","Other.Costs")]
df9$Total.Benefits.M. <- as.numeric(str_replace_all(df9$Total.Benefits.M., c("," = "")))
df9$Other.Costs <- as.numeric(str_replace_all(df9$Other.Costs, c("," = "")))
df9 <- df9 %>%
  dplyr::rename("Total Benefits" = "Total.Benefits.M.",
                "Other Costs" = "Other.Costs")


valid = c("Total Benefits", "Other Costs")
# source: https://fns-prod.azureedge.net/sites/default/files/snap/FY16-State-Activity-Report.pdf
df9$'Household Errors' <- 421.934288
df9$'Agency Errors' <- 188.859846
df9$'Fraud Claims' <- 73.403758

df9 <- as.data.frame(t(df9))
colnames(df9) <- "Cost_millions"
df9$category <- rep("Fraud", length(df9$Cost_millions))
df9$Cause <- rownames(df9)


rownames(df9) <- NULL

for (i in 1:5){
    if (df9$Cause[i] %in% valid){
      df9$category[i] <- "Valid"
}
}



# 8 - Total cost of SNAP vs. Fraud costs, 2016
ggplot(data=df9, aes(x=category, y=Cost_millions, group=Cause, label=Cost_millions)) +
  geom_col(aes(fill=Cause)) +
  theme_bill() +
  xlab("Total Costs vs. Fraud") +
  ylab("Cost (Millions)") +
  ggtitle("SNAP Program Costs vs. Fraud Expenses - 2016", subtitle = "Sources: US Census Bureau, USDA, fns-prod.azureedge.net") +
  scale_y_continuous(labels=scales::dollar_format()) +
  theme(legend.position = "bottom") +
  coord_flip() +
  theme(legend.title = element_text(size = 10),
        legend.text = element_text(size = 8))
```
<br><br>
Despite anti-welfare advocates insisting on punishing welfare recipients due to the abusers and defrauders or imposing harsh restrictions that cut off recipients, fraud from participants makes up a tiny amount of the total money spent annually on SNAP - nearly all of which goes towards food or overhead costs. Household errors made up most fraud costs in 2016 at $424 million. Agency errors cost the program $188 million and fraudulent claims cost $73 million - compared to a total program expenditure of $70.9 billion.
<br><br><br>
<b>9.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# source: https://fred.stlouisfed.org/series/UNRATE
df10 <- read.csv("https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23e1e9f0&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=1168&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=UNRATE&scale=left&cosd=1948-01-01&coed=2021-04-01&line_color=%234572a7&link_values=false&line_style=solid&mark_type=none&mw=3&lw=2&ost=-99999&oet=99999&mma=0&fml=a&fq=Annual&fam=avg&fgst=lin&fgsnd=2020-02-01&line_index=1&transformation=lin&vintage_date=2021-05-17&revision_date=2021-05-17&nd=1948-01-01")
df10$DATE <- substring(df10$DATE, 1, 4)
df10 <- df10 %>%
  dplyr::rename(Year=DATE)
df10 <- as.data.frame(inner_join(df8.1, df10, by="Year"))
df10$UNRATE <- as.numeric(as.character(df10$UNRATE)) / 100
df10$Year <- as.numeric(as.character(df10$Year))


colors <- c("Unemployment Rate" = "darkblue", "Low Food Security" = "red", "Very Low Food Security" = "maroon")



# 9 - hunger vs. unemployment
ggplot() +
  geom_line(data=df10, aes(x=Year, y=UNRATE, color="Unemployment Rate"), size=2) +
  geom_line(data=df10, aes(x=Year, y=Food_Insecure, color="Low Food Security"), size=2) +
  geom_line(data=df10, aes(x=Year, y=Very_Insecure, color="Very Low Food Security"), size=2) +
  labs(y="", color = "Legend:", caption="- Low Food Security includes Very Low Food Security.", size = 2.5, colour = "black") +
  scale_color_manual(values = colors) +
  theme_bill() +
  scale_y_continuous(labels=scales::percent_format(), limits=c(0, 0.15)) +
  ggtitle("Hunger vs. Unemployment Rate, 1995-2019", subtitle="Sources: US Census Bureau, USDA, fred.stlouisfed.org")

```
<br><br>
The purpose of this line graph is to illustrate the effect of unemployment on hunger over the past 25 years. The spike in unemployment to nearly 10% in 2009-2010 was predicated with a large increase in hunger in 2008 - possibly explained by home foreclosures from the Subprime Mortgage Crisis. While unemployment steadily decreased to below 5% by 2015, the percentage of Americans suffering from hunger didn't return to pre-Recession numbers until the end of the 2010s.
<br><br><br>
<b>10.)</b>

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# source: https://data.worldbank.org/indicator/SP.POP.GROW
df11 <- read_excel("C:/Users/Bill's PC/Desktop/School Work/PPOL 646-3/final/data/API_SP.POP.GROW_DS2_en_excel_v2_2253989.xls", skip=2, col_names = TRUE)
df11 <- as.data.frame(df11)
df11 <- df11 %>%
  dplyr::filter(df11$'Country Code' == "USA")
df11 <- t(df11)
df11 <- na.omit(df11)
colnames(df11) <- 'PopGrowthRate'
df11 <- df11[5:length(df11)]
df11$PopGrowthRat <- as.numeric(df11) / 100
df11$Year <- rep(0, length(df11$PopGrowthRat))

for (i in 1961:2019){
  df11$Year[i-1960] <- i
}

df11 <- as.data.frame(cbind(df11$Year, df11$PopGrowthRat))
colnames(df11) <- c("Year","PopGrowthRate")


colors2 <- c("Population Growth" = "darkgreen", "Low Food Security" = "red")



# 10 - Population growth vs. hunger
ggplot() +
  geom_line(data=df10, aes(x=Year, y=Food_Insecure, color="Low Food Security"), size=2) +
  geom_line(data=df11, aes(x=Year, y=PopGrowthRate, color="Population Growth"), size=1) +
  labs(y="", color = "Legend:") +
  scale_color_manual(values = colors2) +
  theme_bill() +
  scale_x_continuous(limits=c(1994, 2020)) +
  scale_y_continuous(labels=scales::percent_format(), limits=c(0, 0.15)) +
  ggtitle("Population Growth Rate vs. Hunger, 1995-2019", subtitle="Sources: US Census Bureau, USDA, World Bank")

```
<br><br>
This line graph shows the trend for percentages of both low food security and population growth rate. Due to the relative economic hardship the United States has faced since the Great Recession, the resulting rise in hunger and its slow recovery, and the significant composition of children and their mothers, I expected to see a decline in the population growth rate over time if the decrease in birth rate was significant enough to drag it down. While there is no obvious correlation here, hunger and lack of appropriate nutrition to America's mothers and children are a plausible cause for the steady decline in population growth since 1995. Factors such as higher levels of education and wider availability of birth control are also reasonable suggestions for this. In Today's context, President Joe Biden has been posited as an "anti-Reagan" after laying out a massive jobs plan to a Joint Session in late April 2021 to fix the COVID-19 unemployment crisis, and attacking the Trickle Down Economics that have been the standard economic policy since Reagan himself. The effects of this spending will be noteworthy in determining if a return to Big Government can rapidly and measurably reduce economic hardship, and the suffering associated.
<br><br><br>

<br><br><br>

<b><u>Data sources:</u></b> <br>
1. https://www.kaggle.com/jpmiller/publicassistance <br>
2. https://fred.stlouisfed.org/graph/fredgraph.csv?bgcolor=%23e1e9f0&chart_type=line&drp=0&fo=open%20sans&graph_bgcolor=%23ffffff&height=450&mode=fred&recession_bars=on&txtcolor=%23444444&ts=12&tts=12&width=748&nt=0&thu=0&trc=0&show_legend=yes&show_axis_titles=yes&show_tooltip=yes&id=FPCPITOTLZGUSA&scale=left&cosd=1960-01-01&coed=2019-01-01&line_color=%234572a7&link_values=false&line_style=solid&mark_type=none&mw=3&lw=2&ost=-99999&oet=99999&mma=0&fml=a&fq=Annual&fam=avg&fgst=lin&fgsnd=2009-06-01&line_index=1&transformation=lin&vintage_date=2021-04-28&revision_date=2021-04-28&nd=1960-01-01 <br>
3. https://www2.census.gov/programs-surveys/popest/datasets/2010-2019/counties/totals/co-est2019-alldata.csv <br>
4. https://datacenter.prb.org/download/usdata/indicator/gini/csv <br>
5. http://navajotimes.com/news/2012/0112/012612census.php <br>
6. https://www.census.gov/newsroom/facts-for-features/2017/aian-month.html#:~:text=6.7%20million,the%20total%20population%20in%202016. <br>
7. https://www.ncsl.org/Portals/1/Documents/magazine/articles/2018/SL_0118-Stats.pdf <br>
8. https://community-nutrition-education.extension.org/wp-content/uploads/2020/09/MPR-Impact-Report_Print-Ready.pdf <br>
9. https://www.ers.usda.gov/media/10674/trends.xlsx <br>
10. https://fns-prod.azureedge.net/sites/default/files/snap/FY16-State-Activity-Report.pdf <br>
11. https://fred.stlouisfed.org/series/UNRATE <br>
12. https://data.worldbank.org/indicator/SP.POP.GROW <br>


<br><br>

<b><u>Informational sources:</u></b> <br>
1.  Borrelli, Christopher. “Reagan Used Her, the Country Hated Her. Decades Later, the Welfare Queen of Chicago Refuses to Go Away.” <i>Chicagotribune.com</i>, 10 June 2019, www.chicagotribune.com/entertainment/ct-ent-welfare-queen-josh-levin-0610-story.html.<br>
2.  Bryant, Nick. “Opinion | Biden Isn't FDR. He's the Anti-Reagan.” <i>The Washington Post</i>, WP Company, 22 Apr. 2021, www.washingtonpost.com/opinions/2021/04/22/biden-isnt-fdr-hes-anti-reagan/.

<br><br>
